---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EJSCREENbatch

<!-- badges: start -->
<!-- badges: end -->

To streamline initial EJ analysis efforts over *multiple locations of interest*, the EJSCREENbatch R package was developed to leverage the national demographic and environmental datasets made available through the U.S. Environmental Protection Agency. 

This package offers canned functions to extract and process Census block group EJSCREEN data for buffered areas surrounding user-provided [sf](https://r-spatial.github.io/sf/) geographies. Users can request different buffer distances or GIS inclusion methods. Moreover, buffers can be distance-based over land (i.e. drawing circles around lat/longs) or water (leveraging the [NHDPlus network](https://usgs-r.github.io/nhdplusTools/) to measure up/downstream).

A number of standardized, secondary data visualization functions are also available: correlation plots, heat tables, ranking tables, and maps are among current offerings.

## Installation

You can install the *development* version of EJSCREENbatch from GitHub with the following:

```{r, eval = F}
# requires devtools to install
install.packages('devtools')
library(devtools)

# install from repository
install_github('USEPA/EJSCREENbatch')
library(EJSCREENbatch)
```

To install from GitHub with package vignettes:
```{r, eval = F}
library(devtools)
install_github('USEPA/EJSCREENbatch', build_vignettes=TRUE)
library(EJSCREENbatch)
```

## Example

An overview of the p ackage, [example use of its functions](url.here), and additional documentation are available in the vignette directory.

## EPA Disclaimer

The United States Environmental Protection Agency (EPA) GitHub project code is provided on an "as is" basis and the user assumes responsibility for its use. EPA has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by EPA. The EPA seal and logo shall not be used in any manner to imply endorsement of any commercial product or activity by EPA or the United States Government.