---
title: "An introduction to the EJSCREENbatch tool"
author: "A.R. El-Khattabi, Morgan Teachey & Adam Theising"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{EJSCREENbatch introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Environmental justice (EJ) analyses summarize demographics of local populations and environmental burden in order to study the impacts policy decisions may have on different communities. While impacts are often localized, it is helpful for decision makers to be able to assess impacts across many areas at once, allowing for comparison between communities and across regulatory actions. To streamline initial screening-level EJ analysis efforts, this package was developed to leverage national demographic and environmental datasets made available through the U.S. Environmental Protection Agency. Specifically, it allows users to analyze EJ summary statistics for an unlimited number of locations (coordinates or water features) and produces customized figures and maps based on user specifications.

The EJSCREENbatch R package primarily relies on data provided by [EJSCREEN](https://www.epa.gov/ejscreen), a mapping and screening tool maintained by EPA that provides demographic and environmental impact data on the Census Block Group (CBG) level for the United States. For information on how these data were prepared, refer to the [EJSCREEN Technical Information Guidance](https://www.epa.gov/ejscreen/technical-documentation-ejscreen)

 

## Getting started
To access the latest version of the package, simply install from the EPA's Github Repo:
```{r, eval = F}
# Import EJSCREENbatch package 
library(devtools)
install_github(repo = "USEPA/EJSCREENbatch")
```


Then load the package, and set your working directory to that of your input data. (Note: as of V1.0, you must also create an object with the filepath housing the SEDAC raster.)
```{r, warning = F, message = F}
library(EJSCREENbatch)

# Set directory and local pointer to SEDAC raster file.
  user<-tolower(Sys.getenv("USERNAME"))

  if(user=="aelkhatt"){
    setwd("C:/Users/aelkhatt/OneDrive - Environmental Protection Agency (EPA)/Documents/projects/environmental justice/EJSCREEN")
    path.raster.layer <- "C:/Users/aelkhatt/OneDrive - Environmental Protection Agency (EPA)/Documents/projects/environmental justice/EJScreen/US Census Grid_SF2010_TIFF"

  } else if(user=="atheisin"){
    setwd("C:/Users/atheisin/OneDrive - Environmental Protection Agency (EPA)/ejscreen")
    raster.path <- 'C:/Users/atheisin/Downloads/EJ_example/US Census Grid_SF2010_TIFF'
    
  } else if(user=="mteachey"){
    print("Morgan")
    
  }
  
```
  
## Input data requirements

First, the user supplies input location data. Since this is a batch tool, the data can (and should!) include several input locations. These locations could be points in space (i.e. lat/long coordinates of an emitting facility), boundary shapes (i.e. a set of polygons representing wetlands or lagoons), or a waterbody identifier from the NHDPlus database (more on this below).

Here create a toy data set by drawing 20 random latitude/longitude coordinates from the continental USA. These points will serve as our "locations of interest" for the vignette's sample EJ analysis.

```{r}
## Draw random points from the lower 48
# Bounding box is: Extent: (-124.848974, 24.396308) * (-66.885444, 49.384358)
set.seed(202)
rnd.pts <- as.data.frame(cbind(name = paste0('facility_name_',1:20),
                 longitude = rnorm(20)*8.75 + -95.15, 
                 latitude = rnorm(20)*3 + 37)) %>%
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)
```

One important technical item to note: location inputs must be fed into package functions either (1) as a _simple feature_ (sf) dataframe for point/line/polygon shapes or (2) as a _list_ of catchment common identifiers (COMIDs). See the [sf package](https://r-spatial.github.io/sf/index.html) documentation for a primer on using spatial data in R and the [nhdplusTools package](https://usgs-r.github.io/nhdplusTools/) for an overview of the catchment ID data structure.

For an illustrative sense of the geographies that will be screened, we map our location inputs below. In the next section, we draw spatial buffers around our locations to extract demographic and environmental data from EJSCREEN's national database.

```{r, warning = F, message = F}
# See where these imaginary facilities fall:
library(maps); library(ggplot2)
state.data <- map_data('state')
ggplot() + 
  geom_polygon(data=state.data, aes(x=long, y=lat, group = group),
               color="black", fill=NA) +
  geom_sf(data = rnd.pts, color = 'red') +
  theme_minimal()
```

## Analysis with coordinate locations
The locations of interest are then supplied to the EJfunction, which prepares the data for use in all other functions. Since providing a comparison group is essential in all EJ analyses, the data population is reported in percentiles, either within the state a CBG is located in or on the national level, based on user input. The user is able to specify additional preferences, such as:

1.	Buffer distance from which to select proximal CBGs and buffering method using the buff_dist input. 
2.	Buffer GIS option to use to pull data from selected CBGs via the gis_option input. The default is “intersection”. 
3.	Threshold for identifying elevated indicated and potential populations of concern. The default is the 80th percentile, as recommended by EJSCREEN [cite documentation]. 
4.	State, if the user wants to restrict the analysis to a single state. 

This function also creates some initial plots for examination of trends in environmental and demographic indicators: 

```{r}
knitr::knit_exit()
```
```{r}
my.EJ.data <- EJfunction(data_type = "landbased",
           facility_data = facilities, 
           gis_option = "centroid", 
           buff_dist = 1,
           threshold = 80,   
           state = NULL,
           raster.data = path.raster.layer)
```

    ##I want to display a box plot and a correlation plot above in 2 panels, how? would          ##probably need to change the script to print them directly

Boxplots displaying the range of percentiles across all CBGs are automatically generated for demographic and environmental indices, as well as two-way correlation plots, helping users get an idea of which indices might overlap with one another. 


##Comparing index percentiles

It may be useful to know locations of interest or CBGs with multiple indicators above the 80th percentile threshold for follow up analysis. This can be done using the EJRankings function, where the user selects how many results to return.

```{r}
EJRanking(input_data = my.EJ.data,
          rank_type = 'location',
          geography_type = 'US',
          rank_count = 10)
```

These rankings can be used to inform later outputs, such as heat tables. Heat tables reporting the median values for each index can be produced to summarize either all areas of interest, areas with the highest cumulative index values, or a single area. To allow users to visually assess percentiles, values over the 80th, 90th, and 95th percentiles are highlighted in yellow, orange, and red, respectively. 


```{r, fig.show='hold'}
industry.htable <- EJHeatTables(input.data = my.EJ.data,
                                type = "all")
  ##Notes:
    ## Error in .subset(x, j) : invalid subscript type 'list' 

top10.htable <- EJHeatTables(input.data = my.EJ.data,
                             type = "topn",
                             topN = 10,
                             save.option = T)
  ##Notes:
    ##Error in `[.data.frame`(x, i) : object 'geography' not found

facility.htable <- EJHeatTables(input.data = my.EJ.data,
                                type = "single",
                                keepid = 1)
    ##Error: " no applicable method for 'select' applied to an object of class "NULL" "

```

The package also allows users to explore the extent to which indices above the threshold percentile overlap with others. The EJCountTable function produces two-way table that summarizes the number of locations that are within a distance of census blocks above the 80th percentile threshold for environmental and/or demographic indicators. 

```{r}
EJCountTable(my.EJ.data)

```

Users can also view these data by using the EJMaps function, which produces an interactive map of the U.S. that displays all locations of interest as points. Points are shaded based on the number of indicators above the threshold for proximal CBGs, which allows users to easily identify locations that may be good candidates for outreach efforts and further analysis. 

library(leaflet)
library(RColorBrewer)
library(mapview)
```{r}
EJMaps(input.data = my.EJ.data, 
       indic.option = 'total')
```

Maps can be customized to shade locations by the number of environmental and/or demographic indicators above the threshold value. 

```{r}
EJMaps(input.data = my.EJ.data, 
       indic.option = 'environmental')
```


##Water-based EJ screening analysis

All of the above batch analyses can also be completed for water-based features. This package leverages the surface water network established in the National Hydrology Dataset [link to NHD hydrology explanation here]… Users can supply a list catchments or points, and the tool will identify catchments specified distance up or downstream of these locations. Then, the environmental and demographic data for the CBGs falling within or intersecting the buffer distance from the selected water features are summarized as before. 

library(nhdplusTools)
library(arcpullr)

```{r}
my.EJ.waterdata <- EJfunction(data_type="waterbased", 
                              facility_data=facilities[1:10,], 
                              input.type = 'sf', 
                              gis_option = "centroid", 
                              buff_dist = 1, 
                              attains = T)

EJfunction(input.type = "waterbased",
           facility_data = my.EJ.data, 
           input.type = 'catchment',
           gis_option = "intersection", 
           buff_dist = 1,
           threshold = 80,   
           state = NULL,
           ds_mode = "DM",
           attains = T)
```











## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
